\ttitle

### 2014-01-15 Chapters/Chapter1.tex
Finished feedback Chapter 1

In this thesis the goal is to find the temporal segmentation of time series data explicitly, prior to the classification of the activities.

This is done under the assumption that it could be beneficial for classification methods to possess this explicit segmentation, since the model construction phase can use more data than the time-windowed approach would allow.

To find a temporal segmentation, we rely on change detection in time series.
In the context of this research, these time series consists of recordings from inertial sensors found in smartphones, such as the accelerometer signal, during the performance of human activities.
We are interested in real-world data from both in- and outdoor activities in an uncontrolled environment and in a continuous manner.
Although the number of performed activities will be limited, we assume the temporal segmentation must be more robust than in the case of data from a controlled environment.
For the change detection algorithm we will use a special form of a \gls{svm}, the \gls{occ} based \gls{svdd}, as introduced by Tax~\cite{tax2001one}.
This method models the data in the shape of a high-dimensional hypersphere.
From this hypersphere the radius is extracted, which can be used as an indication of change and be processed by one-dimensional change detection algorithms.
The radius of the hypersphere is used in the following relation.
When the performed activity changes, so will the data distribution and characteristics.
This results in a change of the constructed model, especially in the radius of the hypersphere.
Thus, a change in the radius indicates a changes in the performed activity.

\section{Scope}
The scope of this thesis is to find a temporal segmentation of time series data from inertial sensors embedded in smartphones, obtained during the performance of human activities.
As mentioned above, the segmentation is often obtained as a by-product from direct windowed classification.
Other approaches have a very rigid form of segmentation, \eg by limiting the number of possible segment types \cite{himberg2001time,chamroukhi2013joint}, or have applied it to different type of time series data \cite{li2007segmentation,fuchs2010online,guenterberg2009automatic}.
Others have extracted features from windows of data which are used for the segmentation \cite{guo2012adaptive}.
These activities include sitting, standing, walking, running, and ascending and descending staircases.
The problem of finding a temporal segmentation can be seen as a sub-problem to the problem of Activity Classification.
This relation is illustrated in \Cref{fig:thesis_goal}.
The figure visualizes that we can apply classification on time series data.
In many approaches it is applied to (partially overlapping) windows of data.
In the context of \gls{har}, the window length commonly consists of around two seconds of data.
In our setup we perform temporal segmentation, instead of using a fixed window length.
Each segment will consist of homogeneous data, in the sense that the underlying data generating activity should not change within a segment.
Likewise, the previous and next segments (if present) should be generated by a different activity.

@@

To find these segments of homogeneous data, we employ a model construction phase.
The \gls{oc-svm} model transforms the data points to a high-dimensional hyperspherical shape from which we extract the radius as characterizing metric.
The model processes the data in windows of fixed length.
A change in data characteristics should then result in a change of the hypersphere's radius.
Such a (sudden) change in the radius thus reflects a change in the time series data.
It indicates the end of the current and the beginning of a new segment.
In \Cref{sec:literature_review_temporal_segmentation} a detailed overview of temporal segmentation methods is given.
Information about segments can be used in the classification phase of \gls{har}.
Instead of using relatively small windows, it can apply classification to the full segments of data.
The final result, which is outside the scope is this thesis, will be a full classification of activities over the time series data.

\section{One-Class Support Vector Machines}
In the setup, as described in the previous section and visualized in \Cref{fig:thesis_goal}, we employ a model construction phase.
For this model we will use an \acrlong{oc-svm} implementation: the \acrlong{svdd} algorithm by Tax~\cite{tax2001one}.
In earlier research multi-class \glspl{svm} are used for direct classification of inertial sensor data \cite{he2008activity,mountrakis2011support,anguita2012human}.
Others have used \gls{oc-svm} classifiers for novelty or outlier detection \cite{scholkopf1999support,camci2010change,li2003improving,ma2003time,tax1999support}.
The method by Yin \etal  \cite{yin2008sensor} creates a \gls{oc-svm} classifier to filter normal activity traces, using training data.
However, to the best of our knowledge, \gls{oc-svm} methods have not yet been applied to inertial sensor time series data directly to find changes in behavior and create a temporal segmentation.

@@

\section{Data sets}
In recent research two data sets for \gls{har} have been made public.
In order to benchmark our proposed method we applied our method to these data sets.
However, it turned out the data sets were not appropriate for continuous temporal segmentation.
In the \gls{wisdm} set \cite{kwapisz2011activity} the activities are not performed continuously.
There is no transition period between activities.
Since we are interested in the detection changes between consecutive activities, we can not use this data set.
The \gls{uci-har} from \cite{anguita2012human} also lacks continuous recordings.
Furthermore, it seems to have incorrect labels.
Finally, for both data sets there are no visual images of the performed activities.
So in case of ambiguity there is no ground truth to consult.
To overcome these problems, we have created our own real-world data set.
\TODO{Create name for our own data sets}

@@

\section{Contributions}
As a result from our proposed method and application to recorded real-world data sets, this thesis will provide the following contributions:
\begin{enumerate}
  \item \textbf{Continuous recorded human activity data:} the proposed method is applied to our own recorded human activity data.
  This data set can be grounded by using the video recordings of the subject while performing the activities.
  The data set, consisting of raw inertial sensor recordings, manual change point and activity labeling, and video recordings, is used for testing the method and is made available to the public \cite{vlasveld2013continuous}.
  \item \textbf{Application of \gls{oc-svm} methods to inertial sensor data:} to the best of our knowledge, this thesis is the first application of \gls{oc-svm} methods, especially the \gls{svdd} method, to inertial sensor signals in the context of temporal segmentation.
  Furthermore, this thesis contributes to research that has temporal segmentation of inertial sensor data as its primary objective.
  \item \textbf{Adaptation of \acrshort{svcpd}:} we have adopted and simplified the \acrshort{svcpd} method by Camci~\cite{camci2010change} in order to find change points in time series data.
  In our simplification we have not encountered a noticeably decrease in performance.
\end{enumerate}

\section{Thesis structure}
The structure of this thesis is as follows.
In \Cref{Chapter2} a literature review is provided.
We will look at the different interpretations and implementations for change, novelty, and outlier detection.
Previous work in the field of \gls{har} is discussed and we end with existing applications of \glspl{svm} to change detection.
\Cref{Chapter3} further analyses \glspl{svm} and two implementations of \glspl{oc-svm}.
It relates the properties of the constructed \gls{oc-svm} models to change detection in time series data.
That relation is applied to our problem formulation in \Cref{Chapter4}, where we construct our change detection method \TODO{add name}, based on the work of Camci~\cite{camci2010change}.
In \Cref{Chapter5} we apply the method to artificial data and show the objective performance.
In \Cref{Chapter6} the method is applied to our real-world data set \TODO{Add name}.
In that chapter we show the ability of detecting change in \gls{har} data, recorded by inertial sensors.
This thesis is concluded in \Cref{Chapter7}, in which we reflect on the performed research and state possibilities for feature research.

>>>>

The goal of this thesis is to create an method for online temporal segmentation of time series data.
This will be performed explicitly, prior to the classification of data segments to learned activities.

Our assumption is that a temporal segmentation of time series can be beneficial information for classification methods.
Instead of using a fixed window approach, the classification methods can use a larger set of homogeneous data.
Thereby, the methods have more information to create a solid model of the activities.

To find a temporal segmentation, we rely on change detection in time series.
In the context of this research the time series consists of recordings from inertial sensors during the performance of human activities.
The sensors, such as the accelerometer, are nowadays widely available in smartphones.
We are interested in real-world data from both in- and outdoor activities in an unknown environment and in a continuous manner.

Compared to recordings in a controlled laboratory environment, we expect the real-world data to be prone to unexpected characteristics.
The constructed method must be robust in order to handle these.

For the change detection algorithm we will use a special form of a \gls{svm}, the \gls{occ} based \gls{svdd}, as introduced by Tax~\cite{tax2001one}.
This method models the data in the shape of a high-dimensional hypersphere.
From this hypersphere the radius is extracted, which can be used as an indication of change and be processed by one-dimensional change detection algorithms.
The radius of the hypersphere is used in the following relation.
When the performed activity changes, so will the data distribution and characteristics.
This results in a change of the constructed model, especially in the radius of the hypersphere.
Thus, a change in the radius indicates a changes in the performed activity.

The remainder of this introduction is organized as follows.
In \Cref{sec:intro_scope} the scope of this thesis is discussed.
A short discussing on the workings of an \acrlong{oc-svm} is provided in \Cref{sec:intro_ocsvm}.
The data sets we have used in this work are discussed in \Cref{sec:intro_data_sets}.
\Cref{sec:intro_contributions} gives a short overview of our contributions.
Finally, \Cref{sec:intro_thesis_structure} outlines the structure of this thesis.

\section{Scope}\label{sec:intro_scope}
The scope of this thesis are segmentation methods of inertial sensor data in the field of machine learning.
The type of data is inertial sensor time series, collected by on-body worn sensors as found in smartphones.
These time series are recorded during the performance of real-world human activities.
The setting of the activities are real-world unknown environments.

As mentioned in the previous section, a segmentation is often obtained as a by-product from direct windowed classification.
Other approaches have a very rigid form of segmentation, \eg by limiting the number of possible segment types \cite{himberg2001time,chamroukhi2013joint}, or have applied it to different type of time series data.
In Fuchs \etal \cite{fuchs2010online} artificial sinusoidal data is processed.
Li and Zheng \cite{li2007segmentation} use visual motion data streams to find patterns of activities.
The method of Guenterberg \etal \cite{guenterberg2009automatic} is able to distinguish periods of rest and activity in inertial sensor data streams.
Others have extracted features from windows of data which are used for the segmentation \cite{guo2012adaptive}.
These activities include sitting, standing, walking, running, and ascending and descending staircases.

The problem of finding a temporal segmentation can be seen as a sub-problem to the problem of Activity Classification.
This relation is illustrated in \Cref{fig:thesis_goal}.
The figure visualizes that we can apply classification on time series data.
In many approaches it is applied to (partially overlapping) windows of data.
In the context of \gls{har}, the window length commonly consists of around two seconds of data.
In our setup we perform temporal segmentation, instead of using a fixed window length.
Each segment will consist of homogeneous data, in the sense that the underlying data generating activity should not change within a segment.
Likewise, the previous and next segments (if present) should be generated by a different activity.

@@

To find these segments of homogeneous data, we employ a model construction phase.
The \acrlong{oc-svm} transforms the data to a higher dimensional space.
Non-linear mapping functions are used to add flexibility to the classification decision boundary.
In the higher dimensional space an enclosing boundary around the data points, in the shape of a hypersphere, is constructed.
The size, expressed as the radius, of the hypersphere depends on the distribution of the data points in the model.
We will use the radius of this hypersphere as a characterizing metric and base our change detection on it.

The model processes the data in windows of fixed length.
A change in data distribution and characteristics should then result in a change of the hypersphere's radius.
Such a (sudden) change in the radius thus reflects a change in the time series data.
It indicates the end of the current and the beginning of a new segment.
In \Cref{sec:literature_review_temporal_segmentation} a detailed overview of temporal segmentation methods is given.
Information about segments can be used in the classification phase of \gls{har}.
Instead of using relatively small windows, it can apply classification to the full segments of data.
The final result, which is outside the scope is this thesis, will be a full classification of activities over the time series data.

\section{One-Class Support Vector Machines}\label{sec:intro_ocsvm}
In the setup, as described in the previous section and visualized in \Cref{fig:thesis_goal}, we employ a model construction phase.
For this model we will use an \acrlong{oc-svm} implementation: the \acrlong{svdd} algorithm by Tax~\cite{tax2001one}.
In earlier research multi-class \glspl{svm} are used for direct classification of inertial sensor data \cite{he2008activity,mountrakis2011support,anguita2012human}.
Others have used \gls{oc-svm} classifiers for novelty or outlier detection \cite{scholkopf1999support,camci2010change,li2003improving,ma2003time,tax1999support}.
The method by Yin \etal  \cite{yin2008sensor} creates a \gls{oc-svm} classifier to filter normal activity traces, using training data.
However, to the best of our knowledge, \gls{oc-svm} methods have not yet been applied to inertial sensor time series data directly to find changes in behavior and create a temporal segmentation.

@@

\section{Data sets}\label{sec:intro_data_sets}
In recent research two data sets for \gls{har} have been made public.
In order to benchmark our proposed method we applied our method to these data sets.
However, it turned out the data sets were not appropriate for continuous temporal segmentation.
In the \gls{wisdm} set \cite{kwapisz2011activity} the activities are not performed continuously.
There is no transition period between activities.
Since we are interested in the detection changes between consecutive activities, we can not use this data set.
The \gls{uci-har} from \cite{anguita2012human} also lacks continuous recordings.
Furthermore, it seems to have incorrect labels.
Finally, for both data sets there are no visual images of the performed activities.
So in case of ambiguity there is no ground truth to consult.
To overcome these problems, we have created our own real-world data set: \acrlong{almende-data} \cite{vlasveld2014acras}.

@@

\section{Contributions}\label{sec:intro_contributions}
As a result from our proposed method and application to recorded real-world data sets, this thesis will provide the following contributions:
\begin{enumerate}
  \item \textbf{Continuous recorded human activity data:} the proposed method is applied to our own recorded human activity data.
  This data set can be grounded by using the video recordings of the subject while performing the activities.
  The data set, consisting of raw inertial sensor recordings, manual change point and activity labeling, and video recordings, is used for testing the method and is made available to the public \cite{vlasveld2014acras}.
  \item \textbf{Application of \gls{oc-svm} methods to inertial sensor data:} to the best of our knowledge, this thesis is the first application of \gls{oc-svm} methods, especially the \gls{svdd} method, to inertial sensor signals in the context of temporal segmentation.
  Furthermore, this thesis contributes to research that has temporal segmentation of inertial sensor data as its primary objective.
  \item \textbf{Adaptation of \acrshort{svcpd}:} we have adopted and simplified the \acrshort{svcpd} method by Camci~\cite{camci2010change} in order to find change points in time series data.
  In our simplification we have not encountered a noticeably decrease in performance.
\end{enumerate}

\section{Thesis structure}\label{sec:intro_thesis_structure}
The structure of this thesis is as follows.
In \Cref{Chapter2} a literature review is provided.
We will look at the different interpretations and implementations for change, novelty, and outlier detection.
Previous work in the field of \gls{har} is discussed and we end with existing applications of \glspl{svm} to change detection.
\Cref{Chapter3} further analyses \glspl{svm} and two implementations of \glspl{oc-svm}.
It relates the properties of the constructed \gls{oc-svm} models to change detection in time series data.
That relation is applied to our problem formulation in \Cref{Chapter4}, where we construct our change detection method \TODO{add name}, based on the work of Camci~\cite{camci2010change}.
In \Cref{Chapter5} we apply the method to artificial data and show the objective performance.
In \Cref{Chapter6} the method is applied to our real-world \gls{almende-data} \cite{vlasveld2014acras}.
In that chapter we show the ability of detecting change in \gls{har} data, recorded by inertial sensors.
This thesis is concluded in \Cref{Chapter7}, in which we reflect on the performed research and state possibilities for feature research.

### 2014-01-15 Chapters/Chapter6/data_sets.tex
Finished feedback Chapter 1

Using the video recordings, the data sets are manually annotated, as shown in the plots of \Cref{fig:plots_subject_1,fig:plots_subject_2,fig:plots_subject_3}.
The graphs display each from a single recording the inertial sensor values (see \Cref{tab:recorded_metrics} for explanation of the metrics), with manually annotated activities above the data.
The manually determined change points are used to draw quality conclusion over the algorithm's performance, by comparing it to the discovered change points.
All of the used data sets, including the video material and manual labeling, are made available to the public \cite{vlasveld2013continuous} for further research.
\TODO{Publish web page with data sets}
The following section discusses the performed activities during the recordings.

>>>>

Using the video recordings, the data sets are manually annotated, as shown in the plots of \Cref{fig:plots_subject_1,fig:plots_subject_2,fig:plots_subject_3}.
The graphs display each from a single recording the inertial sensor values (see \Cref{tab:recorded_metrics} for explanation of the metrics), with manually annotated activities above the data.
The manually determined change points are used to draw quality conclusion over the algorithm's performance, by comparing it to the discovered change points.
All of the used data sets, including the video material and manual labeling, are made available to the public as \gls{almende-data} \cite{vlasveld2014acras} for further research.
The following section discusses the performed activities during the recordings.

### 2014-01-15 Chapters/Chapter6/results.tex
Finished feedback Chapter 1

\subsection{Subjective observations and remarks}
It can be ambiguous where to place the (manually annotated) change points, in the case of real world data.
The objective quality measure can give a distorted view on the performance.
To overcome this, we will state observations and remarks found after close (visual) inspection of the annotated and discovered change points.
In \Cref{fig:plots_all_runs_results} all the results are plotted.
The black solid lines indicate our manually determined change points.
The dashed purple lines indicate the change points as discovered by our method.
We will start with general observations applicable to all the runs and will continue with individual remarks for each run.
Where possible, a graphical illustration will support our observations.

>>>>

\subsection{Subjective observations and remarks}
It can be ambiguous where to place the (manually annotated) change points, in the case of real-world data.
The objective quality measure can give a distorted view on the performance.
To overcome this, we will state observations and remarks found after close (visual) inspection of the annotated and discovered change points.
In \Cref{fig:plots_all_runs_results} all the results are plotted.
The black solid lines indicate our manually determined change points.
The dashed purple lines indicate the change points as discovered by our method.
We will start with general observations applicable to all the runs and will continue with individual remarks for each run.
Where possible, a graphical illustration will support our observations.

### 2014-01-15 glossary-acronyms.tex
WIP: Working on organization of Chapter 2
WIP: kernels
Add equations
'Refactoring'
Cleanup of sections
WIP
Finished feedback Chapter 1

@@

@@

@@

@@

@@

@@

>>>>

\newacronym{almende-data}{ACRAS}{Almende Continuous Real-world Activities with on-body smartphones Sensors Data Set}

@@

\newacronym{dr}{DR}{Dimensionality Reduction}

\newacronym{far}{FAR}{False Alarm Rate}

@@

\newacronym{uci-har}{UCI HAR}{Human Activity Recognition Using Smartphones Data Set}

@@

\newacronym{plr}{PLR}{Piecewise Linear Representation}

@@

\newacronym{rkhs}{RKHS}{Reproducing Kernel Hilbert Space}

@@

\newacronym{svr}{SVC}{Support Vector Regression}

@@

\newacronym{wisdm}{WISDM}{Wireless Sensor Data Mining}
